---
title: "Advanced Survival Analysis — Rats Dataset"
author: "Walter Nyamutamba, Samuel Agyekum, Honorine Mujawayezu, Belyse Kaneza, Grace Kitonyi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 6
)
```

## Introduction

This report extends the parametric survival analysis from Project 1 by applying
non-parametric and semi-parametric methods to the `rats` dataset. We compare
Kaplan–Meier and Nelson–Aalen estimates, perform formal hypothesis tests, fit
Cox proportional hazards models with full diagnostics, and synthesise findings
across all three analytical frameworks.

---

## Packages and Data

```{r packages}
library(survival)
library(flexsurv)
library(survminer)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(broom)
library(gridExtra)

style_table <- function(df, caption) {
  df %>%
    kbl(caption = caption) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = FALSE
    ) %>%
    row_spec(0, bold = TRUE, background = "#2E86AB", color = "white")
}
```

```{r data-prep}
data(rats)

rats <- rats %>%
  mutate(
    litter    = as.factor(litter),
    event     = as.numeric(status),
    treatment = factor(ifelse(rx == 1, "Treated", "Control"),
                       levels = c("Control", "Treated")),
    sex       = as.factor(sex)
  )

rats_clean <- rats %>%
  select(litter, time, event, treatment, sex) %>%
  na.omit()

surv_obj <- Surv(time = rats_clean$time, event = rats_clean$event)
```
```{r}
rats
```

---

## Step 1: Kaplan–Meier Estimator

### Overall survival curve

```{r km-overall, fig.height=7}
km_overall <- survfit(surv_obj ~ 1, data = rats_clean)

ggsurvplot(
  km_overall,
  data             = rats_clean,
  conf.int         = TRUE,
  risk.table       = TRUE,
  palette          = "#2E86AB",
  ggtheme          = theme_minimal(),
  title            = "Overall Kaplan–Meier Survival Curve",
  xlab             = "Time (days)",
  ylab             = "Survival Probability S(t)",
  risk.table.title = "Number at Risk"
)
```

### Median survival and confidence interval

```{r km-median}
km_tab <- summary(km_overall)$table

data.frame(
  Quantity = c("Median survival", "95% CI Lower", "95% CI Upper", "RMST"),
  Value    = c(
    ifelse(is.na(km_tab["median"]), "Not reached", km_tab["median"]),
    ifelse(is.na(km_tab["0.95LCL"]), "Not reached", km_tab["0.95LCL"]),
    ifelse(is.na(km_tab["0.95UCL"]), "Not reached", km_tab["0.95UCL"]),
    paste(round(km_tab["rmean"], 2), "days")
  )
) %>%
  style_table("Table 1: Overall Median Survival and RMST")
```

**Q1: What is the median survival time?**

The median survival time is not reached in either group because the overall event
rate is only 14% — the KM curve never drops below 0.5, so the median is
mathematically undefined. We report the Restricted Mean Survival Time (RMST)
instead: the average rat remained tumour-free for **`r round(km_tab["rmean"], 2)` days**
out of the 104-day study period.

### Survival probabilities at key time points

```{r km-timepoints}
time_quantiles <- quantile(rats_clean$time, probs = c(0.25, 0.50, 0.75))

km_tp <- summary(km_overall, times = as.numeric(time_quantiles))

data.frame(
  Percentile   = c("25th", "50th", "75th"),
  Time_days    = round(time_quantiles, 1),
  Survival     = round(km_tp$surv,  4),
  Lower_95CI   = round(km_tp$lower, 4),
  Upper_95CI   = round(km_tp$upper, 4),
  N_at_risk    = km_tp$n.risk
) %>%
  style_table("Table 2: Survival Probabilities at Percentile Time Points")
```

**Q2: Survival probabilities at key time points:**

At the 25th percentile of observed time (80.75 days), approximately 97% of
rats are still tumour-free. At the 50th percentile (98 days), about 91%
remain tumour-free. By the 75th percentile (104 days), survival probability
is still above 85%, reflecting the low overall event rate in this study.

### Stratified KM curves by treatment

```{r}
cat("Levels of treatment:", levels(rats_clean$treatment), "\n")
cat("Table of treatment:\n")
print(table(rats_clean$treatment))
cat("Class:", class(rats_clean$treatment), "\n")
```


```{r km-treatment, fig.height=7}
# force remove any existing rats object and reload completely fresh
if (exists("rats")) rm(rats)
data(rats, package = "survival")

# check what rx actually looks like before recoding
cat("rx unique values:", unique(rats$rx), "\n")
cat("rx class:", class(rats$rx), "\n")
cat("rx table:\n"); print(table(rats$rx))

rats <- rats %>%
  mutate(
    litter    = as.factor(litter),
    event     = as.numeric(status),
    # use as.numeric() to force proper comparison regardless of storage type
    treatment = factor(
      ifelse(as.numeric(rx) == 1, "Treated", "Control"),
      levels = c("Control", "Treated")
    ),
    sex = as.factor(sex)
  )

# verify the fix worked
cat("\ntreatment table after recoding:\n")
print(table(rats$treatment))

rats_clean <- rats %>%
  select(litter, time, event, treatment, sex) %>%
  na.omit()

surv_obj <- Surv(time = rats_clean$time, event = rats_clean$event)
```

### Stratified KM curves by sex

```{r km-sex, fig.height=7}
km_sex <- survfit(surv_obj ~ sex, data = rats_clean)

ggsurvplot(
  km_sex,
  data             = rats_clean,
  conf.int         = TRUE,
  risk.table       = TRUE,
  pval             = TRUE,
  pval.method      = TRUE,
  palette          = c("#A23B72", "#3B7080"),
  legend.labs      = c("Female", "Male"),
  legend.title     = "Sex",
  ggtheme          = theme_minimal(),
  title            = "Kaplan–Meier Curves by Sex",
  xlab             = "Time (days)",
  ylab             = "Survival Probability S(t)",
  risk.table.title = "Number at Risk",
  tables.theme     = theme_cleantable()
)
```

### RMST by group

```{r rmst}
# check actual row names first
km_treatment <- survfit(Surv(time, event) ~ treatment, data = rats_clean)
km_overall   <- survfit(Surv(time, event) ~ 1, data = rats_clean)

rmst_groups <- summary(km_treatment)$table
cat("Row names in rmst_groups:\n")
print(rownames(rmst_groups))

# use actual row names dynamically
ctrl_row <- rownames(rmst_groups)[grep("Control", rownames(rmst_groups))]
trt_row  <- rownames(rmst_groups)[grep("Treated", rownames(rmst_groups))]

data.frame(
  Group     = c("Overall", "Control", "Treated"),
  RMST_days = round(c(
    summary(km_overall)$table["rmean"],
    rmst_groups[ctrl_row, "rmean"],
    rmst_groups[trt_row,  "rmean"]
  ), 2),
  SE = round(c(
    summary(km_overall)$table["se(rmean)"],
    rmst_groups[ctrl_row, "se(rmean)"],
    rmst_groups[trt_row,  "se(rmean)"]
  ), 4)
) %>%
  style_table("Table 3: Restricted Mean Survival Time by Group")
```

**Q3: How does survival differ between groups?**

The KM curves show clear separation between groups. The treated group drops
faster, particularly after day 75, consistent with the carcinogenic agent
accelerating tumour onset. The RMST difference is 1.83 days (Control 100.38
vs Treated 98.55). The sex stratification reveals a more striking separation —
female rats develop tumours at a far higher rate than males (p < 0.0001),
with male rats remaining almost entirely tumour-free throughout the study.

---

## Step 2: Nelson–Aalen Estimator

### Cumulative hazard function

```{r na-overall, fig.height=6}
na_overall <- survfit(surv_obj ~ 1, data = rats_clean,
                      type = "fleming-harrington")

ggsurvplot(
  na_overall,
  data    = rats_clean,
  fun     = "cumhaz",
  conf.int = TRUE,
  palette = "#2E86AB",
  ggtheme = theme_minimal(),
  title   = "Nelson–Aalen Cumulative Hazard Estimate",
  xlab    = "Time (days)",
  ylab    = "Cumulative Hazard H(t)"
)
```

```{r na-by-group, fig.height=6}
na_rx <- survfit(surv_obj ~ treatment, data = rats_clean,
                 type = "fleming-harrington")

ggsurvplot(
  na_rx,
  data         = rats_clean,
  fun          = "cumhaz",
  conf.int     = TRUE,
  palette      = c("#2E86AB", "#F18F01"),
  legend.labs  = c("Control", "Treated"),
  legend.title = "Group",
  ggtheme      = theme_minimal(),
  title        = "Cumulative Hazard by Treatment Group",
  xlab         = "Time (days)",
  ylab         = "Cumulative Hazard H(t)"
)
```

**Q1: Shape of the cumulative hazard:**

The cumulative hazard function H(t) is concave upward — it accelerates over
time rather than growing linearly. This means the hazard is not constant. The
curve is flat early in the study and steepens sharply after day 75, indicating
that tumour risk is concentrated in the later part of the follow-up period.

### Smoothed hazard function

```{r hazard-smooth, fig.height=5}
na_data <- data.frame(
  time   = na_overall$time,
  cumhaz = na_overall$cumhaz
)
na_data$delta_H <- c(na_data$cumhaz[1], diff(na_data$cumhaz))
na_data$delta_t <- c(na_data$time[1],   diff(na_data$time))
na_data$hazard  <- na_data$delta_H / pmax(na_data$delta_t, 1)

loess_fit            <- loess(hazard ~ time, data = na_data, span = 0.5)
na_data$smooth_haz   <- pmax(predict(loess_fit), 0)

# Weibull parametric hazard for comparison
weibull_overall <- flexsurvreg(surv_obj ~ 1, data = rats_clean, dist = "weibull")
t_seq           <- seq(1, max(rats_clean$time), length.out = 200)
weibull_h       <- summary(weibull_overall, t = t_seq, type = "hazard")[[1]]

plot(na_data$time, na_data$smooth_haz,
     type = "l", col = "#2E86AB", lwd = 2,
     xlab = "Time (days)", ylab = "h(t)",
     main = "Non-Parametric (Nelson–Aalen) vs Weibull Hazard")
lines(weibull_h$time, weibull_h$est,
      col = "#F18F01", lwd = 2, lty = 2)
legend("topleft",
       legend = c("Nelson–Aalen (smoothed)", "Weibull (parametric)"),
       col = c("#2E86AB", "#F18F01"), lwd = 2, lty = c(1, 2), bty = "n")
grid()
```

**Q2: Hazard trend and comparison with Weibull:**

The smoothed non-parametric hazard is monotonically increasing — confirming
the trend implied by the Weibull shape parameter k = 3.667 from Project 1.
Both estimates agree closely in shape and direction. The Weibull parametric
hazard aligns well with the Nelson–Aalen smoothed estimate, particularly in
the acceleration pattern after day 60. This consistency between the
non-parametric and parametric approaches strongly supports the Weibull model
as an appropriate distributional assumption for this data.

---

## Step 3: Non-Parametric Hypothesis Testing

### Log-rank test

```{r logrank}
lr_treatment <- survdiff(surv_obj ~ treatment, data = rats_clean)
lr_sex       <- survdiff(surv_obj ~ sex,       data = rats_clean)

pval_lr_trt <- 1 - pchisq(lr_treatment$chisq, df = 1)
pval_lr_sex <- 1 - pchisq(lr_sex$chisq,       df = 1)

data.frame(
  Comparison  = c("Treatment (Control vs Treated)", "Sex (Female vs Male)"),
  Chi_squared = round(c(lr_treatment$chisq, lr_sex$chisq), 4),
  df          = 1,
  p_value     = round(c(pval_lr_trt, pval_lr_sex), 4),
  Conclusion  = c(
    ifelse(pval_lr_trt < 0.05, "Significant difference", "No significant difference"),
    ifelse(pval_lr_sex < 0.05, "Significant difference", "No significant difference")
  )
) %>%
  style_table("Table 4: Log-Rank Test Results")
```

### Wilcoxon (Breslow) test

```{r wilcoxon}
wc_treatment <- survdiff(surv_obj ~ treatment, data = rats_clean, rho = 1)
wc_sex       <- survdiff(surv_obj ~ sex,       data = rats_clean, rho = 1)

pval_wc_trt <- 1 - pchisq(wc_treatment$chisq, df = 1)
pval_wc_sex <- 1 - pchisq(wc_sex$chisq,       df = 1)

data.frame(
  Comparison  = c("Treatment (Control vs Treated)", "Sex (Female vs Male)"),
  Chi_squared = round(c(wc_treatment$chisq, wc_sex$chisq), 4),
  df          = 1,
  p_value     = round(c(pval_wc_trt, pval_wc_sex), 4),
  Conclusion  = c(
    ifelse(pval_wc_trt < 0.05, "Significant difference", "No significant difference"),
    ifelse(pval_wc_sex < 0.05, "Significant difference", "No significant difference")
  )
) %>%
  style_table("Table 5: Wilcoxon (Breslow) Test Results")
```

### Comparison of tests

```{r test-comparison}
data.frame(
  Test         = c("Log-Rank", "Wilcoxon"),
  Treatment_p  = round(c(pval_lr_trt, pval_wc_trt), 4),
  Sex_p        = round(c(pval_lr_sex, pval_wc_sex), 4)
) %>%
  style_table("Table 6: Log-Rank vs Wilcoxon p-values")
```

**Q1: Log-rank test result:**

The log-rank test gives p = `r round(pval_lr_trt, 4)` for treatment — there
is a statistically significant difference in survival between Control and
Treated groups. For sex, p < 0.0001, an even stronger result.

**Q2: Wilcoxon test and emphasis on early events:**

The Wilcoxon test weights early time points more heavily. For treatment, both
tests agree (both p < 0.05), though the Wilcoxon p-value is slightly larger.
This suggests the survival difference between treatment groups is more
pronounced at later time points — consistent with the KM curves which separate
mainly after day 75. For sex, both tests strongly agree.

**Q3: Consistency of tests:**

Both tests yield consistent conclusions for both comparisons. Treatment
differences are significant under both (p < 0.05), and sex differences are
highly significant under both (p < 0.001). The slightly larger Wilcoxon p-value
for treatment confirms that the treatment effect is driven by late-occurring
tumours rather than early events — biologically plausible given the mechanism
of carcinogenesis.

---

## Step 4: Cox Proportional Hazards Model

### Univariable Cox models

```{r cox-uni-treatment}
cox_rx  <- coxph(Surv(time, event) ~ treatment, data = rats_clean)
cox_sex <- coxph(Surv(time, event) ~ sex,       data = rats_clean)

bind_rows(
  tidy(cox_rx,  exponentiate = TRUE, conf.int = TRUE),
  tidy(cox_sex, exponentiate = TRUE, conf.int = TRUE)
) %>%
  mutate(across(where(is.numeric), ~round(.x, 3))) %>%
  style_table("Table 7: Univariable Cox Model Results")
```

**Q1: Univariable hazard ratios:**

For treatment: HR = `r round(exp(coef(cox_rx)), 3)`,
95% CI [`r round(exp(confint(cox_rx))[1], 3)`,
`r round(exp(confint(cox_rx))[2], 3)`], p = `r round(summary(cox_rx)$coefficients[,"Pr(>|z|)"], 4)`.
Treated rats have approximately double the hazard of tumour development
compared to controls.

For sex: HR = `r round(exp(coef(cox_sex)), 3)`,
95% CI [`r round(exp(confint(cox_sex))[1], 3)`,
`r round(exp(confint(cox_sex))[2], 3)`], p < 0.001.
Male rats have a 95% lower hazard than females — sex is the dominant predictor.

### Multivariable Cox model

```{r cox-multi}
cox_multi <- coxph(Surv(time, event) ~ treatment + sex, data = rats_clean)

tidy(cox_multi, exponentiate = TRUE, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~round(.x, 3))) %>%
  style_table("Table 8: Multivariable Cox Model (Treatment + Sex)")
```

```{r forest-plot, fig.height=4}
ggforest(cox_multi, data = rats_clean)
```

**Q2: Multivariable model — adjusted effects:**

After adjusting for sex, the treatment HR increases slightly from 2.042 to
2.206, and the p-value strengthens from 0.021 to 0.011. This indicates sex
was acting as a slight confounder — adjusting for it reveals a marginally
stronger treatment effect. Both treatment and sex remain significant in the
multivariable model. Male sex is protective (HR = 0.047), remaining the
strongest predictor in the model.

### Cox model with litter clustering

```{r cox-cluster}
cox_cluster <- coxph(
  Surv(time, event) ~ treatment + sex + cluster(litter),
  data = rats_clean
)

tidy(cox_cluster, exponentiate = TRUE, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~round(.x, 3))) %>%
  style_table("Table 9: Cox Model with Litter-Clustered Standard Errors")
```

### Proportional hazards assumption

```{r ph-test}
ph_test <- cox.zph(cox_multi)

as.data.frame(ph_test$table) %>%
  mutate(Variable = rownames(.), across(where(is.numeric), ~round(.x, 4))) %>%
  select(Variable, everything()) %>%
  style_table("Table 10: Schoenfeld Residuals Test (PH Assumption)")
```

```{r schoenfeld-plots, fig.height=5}
par(mfrow = c(1, 2))
plot(ph_test, var = 1,
     main = "Schoenfeld Residuals — Treatment",
     ylab = "Beta(t)", xlab = "Time (days)", col = "#F18F01", lwd = 2)
abline(h = coef(cox_multi)[1], col = "red", lty = 2, lwd = 2)

plot(ph_test, var = 2,
     main = "Schoenfeld Residuals — Sex",
     ylab = "Beta(t)", xlab = "Time (days)", col = "#A23B72", lwd = 2)
abline(h = coef(cox_multi)[2], col = "red", lty = 2, lwd = 2)
par(mfrow = c(1, 1))
```

**Q3: Proportional hazards assumption:**

The Schoenfeld test gives p = 0.020 for treatment — the PH assumption is
violated. The non-flat pattern in the treatment residual plot confirms the
hazard ratio changes over time. Sex has p = 0.438, so PH holds for sex. The
global test gives p = 0.046, indicating the overall model also has some
time-varying behaviour. This violation motivates the use of extended Cox models
and parametric approaches.

### Influential observations

```{r dfbetas, fig.height=5}
dfb <- residuals(cox_multi, type = "dfbeta")

par(mfrow = c(1, 2))
plot(dfb[, 1], ylab = "dfbeta (treatment)",
     main = "Influence on Treatment Coefficient",
     pch = 16, cex = 0.6, col = "#2E86AB")
abline(h = 0, lty = 2, col = "red")

plot(dfb[, 2], ylab = "dfbeta (sex)",
     main = "Influence on Sex Coefficient",
     pch = 16, cex = 0.6, col = "#A23B72")
abline(h = 0, lty = 2, col = "red")
par(mfrow = c(1, 1))
```

**Q4: Influential observations:**

The dfbeta plots show all observations clustered near zero with no extreme
outliers. No single rat is disproportionately influencing either the treatment
or sex coefficient estimate. The model is stable.

### Martingale and deviance residuals

```{r martingale, fig.height=5}
mart_null <- residuals(
  coxph(Surv(time, event) ~ 1, data = rats_clean),
  type = "martingale"
)

dev_resid <- residuals(cox_multi, type = "deviance")

par(mfrow = c(1, 2))
plot(rats_clean$time, mart_null,
     pch = 16, cex = 0.5, col = "#2E86AB",
     xlab = "Time (days)", ylab = "Martingale Residual",
     main = "Martingale Residuals vs Time")
lines(lowess(rats_clean$time, mart_null), col = "red", lwd = 2)
abline(h = 0, lty = 2)

plot(dev_resid, pch = 16, cex = 0.5, col = "#43A047",
     ylab = "Deviance Residual",
     main = "Deviance Residuals")
abline(h = c(-2, 0, 2), lty = c(2, 1, 2), col = c("red", "black", "red"))
par(mfrow = c(1, 1))
```

**Q5: Linearity and functional form:**

Since both covariates are categorical (treatment and sex), the linearity
assumption in the log-hazard does not apply in the usual sense. The Martingale
residual plot shows a roughly flat lowess line around zero, suggesting no
obvious non-linear time trend beyond what the model already captures. Deviance
residuals are mostly within ±2, confirming no severe outliers.

---

## Step 5: Model Refinement — Addressing PH Violation

Since the PH assumption is violated for treatment (p = 0.020), we apply two
remedies: stratification and time-dependent coefficients.

### Strategy 1 — Stratified Cox model

```{r cox-stratified}
cox_strat <- coxph(
  Surv(time, event) ~ treatment + strata(sex),
  data = rats_clean
)

tidy(cox_strat, exponentiate = TRUE, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~round(.x, 3))) %>%
  style_table("Table 11: Stratified Cox Model (Stratified by Sex)")
```

```{r ph-strat}
ph_strat <- cox.zph(cox_strat)

as.data.frame(ph_strat$table) %>%
  mutate(Variable = rownames(.), across(where(is.numeric), ~round(.x, 4))) %>%
  select(Variable, everything()) %>%
  style_table("Table 12: PH Test on Stratified Model")
```

### Strategy 2 — Time-dependent coefficient

```{r cox-td}
# create a numeric version of treatment for tt()
rats_clean$treatment_num <- as.numeric(rats_clean$treatment == "Treated")

cox_td <- coxph(
  Surv(time, event) ~ treatment + sex + tt(treatment_num),
  tt   = function(x, t, ...) x * log(t + 1),
  data = rats_clean
)

tidy(cox_td, exponentiate = TRUE, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~round(.x, 3))) %>%
  style_table("Table 13: Cox Model with Time-Dependent Treatment Coefficient")

data.frame(
  Model = c("Original multivariable", "Stratified by sex",
            "Time-dependent coefficient"),
  AIC   = round(c(AIC(cox_multi), AIC(cox_strat), AIC(cox_td)), 3),
  Notes = c("PH violated for treatment",
            "Allows different baseline per sex",
            "Allows HR to vary with log(t + 1)")
) %>%
  style_table("Table 14: Model Comparison by AIC")
```

### Model comparison

```{r model-aic}
data.frame(
  Model = c("Original multivariable", "Stratified by sex",
            "Time-dependent coefficient"),
  AIC   = round(c(AIC(cox_multi), AIC(cox_strat), AIC(cox_td)), 3),
  Notes = c("PH violated for treatment",
            "Allows different baseline per sex",
            "Allows HR to vary with log(time)")
) %>%
  style_table("Table 14: Model Comparison by AIC")
```

**Q1: Strategy for PH violation:**

Two strategies were used. The stratified model allows a different baseline
hazard for each sex group while still estimating the treatment HR — this
resolves the PH issue for sex and gives a cleaner treatment estimate.
The time-dependent coefficient model includes `treatment × log(t)` which
directly models how the treatment HR changes over time.

**Q2: Final refined model:**

The stratified Cox model is our preferred refined model. It gives a clean
treatment HR that is now free from the sex-related PH issue. The treatment
effect remains significant after stratification, confirming robustness.
Compared to the original model, AIC is lower for the stratified version,
indicating improved model fit.

**Q3: Comparison with initial model:**

The treatment HR from the stratified model is consistent with the original
multivariable estimate (~2.2), confirming that the PH violation does not
fundamentally change the direction or magnitude of the treatment effect —
it only affects the precision of inference over time.

---

## Step 6: Comprehensive Interpretation and Synthesis

### KM vs best parametric model overlay

```{r km-vs-weibull, fig.height=6}
weibull_fit <- flexsurvreg(surv_obj ~ 1, data = rats_clean, dist = "weibull")
t_plot      <- seq(0.1, max(rats_clean$time), length.out = 300)
wb_surv     <- summary(weibull_fit, t = t_plot, type = "survival")[[1]]

plot(km_overall, col = "black", lwd = 2.5, conf.int = FALSE,
     xlab = "Time (days)", ylab = "S(t)",
     main = "KM Curve vs Best Parametric Model (Weibull)",
     ylim = c(0, 1))
lines(wb_surv$time, wb_surv$est, col = "#F18F01", lwd = 2, lty = 2)
legend("bottomleft",
       legend = c("Kaplan–Meier", "Weibull (parametric)"),
       col    = c("black", "#F18F01"),
       lwd    = 2, lty = c(1, 2), bty = "n")
grid()
```

**Q1: KM vs Weibull:**

The Weibull survival curve closely tracks the KM estimate throughout the
follow-up period. Both curves remain above 0.75 throughout the study,
consistent with the low 14% event rate. The close alignment validates the
Weibull as an appropriate parametric model — it captures the empirical
survival pattern without overfitting. The PP-plot from Project 1 further
confirmed this good agreement.

### Nelson–Aalen hazard vs Weibull hazard

The overlay plot in Step 2 shows that the smoothed Nelson–Aalen hazard and
the Weibull parametric hazard follow the same increasing pattern and
accelerate at a similar rate after day 75.

**Q2: Alignment of hazard estimates:**

The non-parametric hazard trend aligns closely with the Weibull hazard,
with both showing a monotonically increasing pattern consistent with
k = 3.667 > 1. This alignment strongly supports the Weibull distributional
assumption — if the parametric model were misspecified, we would expect
clear divergence between the two hazard estimates.

### Connecting hypothesis tests to Cox model

```{r test-cox-comparison}
data.frame(
  Method       = c("Log-Rank test", "Wilcoxon test",
                   "Univariable Cox", "Multivariable Cox"),
  Treatment_p  = round(c(pval_lr_trt, pval_wc_trt,
                          summary(cox_rx)$coefficients[, "Pr(>|z|)"],
                          summary(cox_multi)$coefficients["treatmentTreated",
                                                          "Pr(>|z|)"]), 4),
  Sex_p        = c(round(pval_lr_sex, 4), round(pval_wc_sex, 4),
                   round(summary(cox_sex)$coefficients[, "Pr(>|z|)"], 4),
                   round(summary(cox_multi)$coefficients["sexm",
                                                         "Pr(>|z|)"], 4))
) %>%
  style_table("Table 15: Consistency of p-values Across Methods")
```

**Q3: Relationship between tests and Cox model:**

All methods give consistent conclusions. The log-rank p-value for treatment
(0.018) closely matches the univariable Cox p-value (0.021) — this is
expected because the log-rank test is equivalent to the score test in a
Cox model with a single binary covariate. The Wilcoxon p-value is slightly
larger for treatment (0.023), reflecting its down-weighting of late events
where the treatment effect is concentrated. For sex, all four methods give
p < 0.001, confirming this is the most robust finding in the analysis.

### Overall survival summary

```{r survival-summary}
data.frame(
  Quantity = c(
    "Total rats", "Events (tumours)", "Censored", "Event rate",
    "RMST — Overall", "RMST — Control", "RMST — Treated",
    "RMST difference",
    "Log-rank p (treatment)", "Log-rank p (sex)",
    "Cox HR — Treated vs Control (adjusted)",
    "Cox HR — Male vs Female (adjusted)",
    "Weibull shape k", "Weibull scale λ",
    "Mean tumour-free survival (overall)",
    "Mean tumour-free survival (Control)",
    "Mean tumour-free survival (Treated)",
    "Frailty variance θ"
  ),
  Value = c(
    300, 42, 258, "14%",
    paste(round(summary(km_overall)$table["rmean"], 2), "days"),
    paste(round(rmst_groups["treatment=Control", "rmean"], 2), "days"),
    paste(round(rmst_groups["treatment=Treated", "rmean"], 2), "days"),
    paste(round(rmst_groups["treatment=Control", "rmean"] -
                  rmst_groups["treatment=Treated", "rmean"], 2), "days"),
    round(pval_lr_trt, 4),
    round(pval_lr_sex, 4),
    paste(round(exp(coef(cox_multi)["treatmentTreated"]), 3),
          "[1.203, 4.044]"),
    paste(round(exp(coef(cox_multi)["sexm"]), 3), "[0.011, 0.193]"),
    "3.667", "160.6",
    "~150.6 days",
    "~173.2 days",
    "~124.8 days",
    "2.02"
  )
) %>%
  style_table("Table 16: Comprehensive Results Summary")
```

**Q4: Overall survival experience:**

The median survival was not reached in either group due to the low event
rate (14%). On average, rats remained tumour-free for approximately 99.8 days
out of the 104-day study period (RMST). Control rats lasted 1.83 days longer
on average than treated rats — a small absolute difference that becomes more
meaningful in context of the doubling of hazard seen in the Cox models.

**Q5: Nature of the hazard:**

The hazard is monotonically increasing over time, confirmed by both the
non-parametric Nelson–Aalen estimate and the Weibull shape parameter k = 3.667.
This means tumour risk accelerates as rats age through the study period —
a rat that has survived 90 days faces substantially higher risk than one that
just entered the study. The increasing hazard also explains why the Exponential
model (which assumes constant hazard) was decisively rejected with ΔAIC = 51.3
in Project 1.

**Q6: Key predictors from the final Cox model:**

Two variables are strong independent predictors:

- **Treatment (HR = 2.21, 95% CI [1.20, 4.04], p = 0.011):** Treated rats
  have more than double the instantaneous hazard of tumour development
  compared to controls at any given time, after adjusting for sex. This
  directly reflects the effect of the carcinogenic agent.

- **Sex (HR = 0.047, 95% CI [0.011, 0.193], p < 0.001):** Male rats have
  a 95.3% lower hazard than female rats. This is the largest effect in the
  analysis and is biologically important — female rats in this study are
  overwhelmingly driving the tumour events (40 out of 42).

**Q7: Clinical and practical implications:**

The carcinogenic drug significantly accelerates tumour onset. The frailty
variance θ = 2.02 confirms that litter identity is a strong determinant of
tumour risk beyond the observed covariates — rats within the same litter share
unobserved genetic susceptibility. Any future study using this carcinogen
should account for litter clustering in its analysis and design. The sex
effect is dramatic enough to suggest that studies of this carcinogen should
either be sex-stratified or restrict to one sex to reduce unexplained
variability. The PH violation for treatment suggests the carcinogen's effect
may wane or strengthen over time — a time-varying effect that future
longitudinal studies should monitor explicitly.

---

## Session Info

```{r session-info}
sessionInfo()
```
```{r}
# create output folders
if (!dir.exists("outputs"))        dir.create("outputs")
if (!dir.exists("outputs/plots"))  dir.create("outputs/plots")
if (!dir.exists("outputs/tables")) dir.create("outputs/tables")

# ── PLOTS ──────────────────────────────────────────────────────────────────

# KM overall
png("outputs/plots/km_overall.png", width = 900, height = 700, res = 120)
print(ggsurvplot(
  km_overall, data = rats_clean, conf.int = TRUE, risk.table = TRUE,
  palette = "#2E86AB", ggtheme = theme_minimal(),
  title = "Overall Kaplan–Meier Survival Curve",
  xlab = "Time (days)", ylab = "S(t)",
  risk.table.title = "Number at Risk"
))
dev.off()

# KM by treatment
png("outputs/plots/km_by_treatment.png", width = 900, height = 700, res = 120)
print(ggsurvplot(
  km_treatment, data = rats_clean, conf.int = TRUE, risk.table = TRUE,
  pval = TRUE, palette = c("#2E86AB", "#F18F01"),
  legend.labs = c("Control", "Treated"), legend.title = "Group",
  ggtheme = theme_minimal(),
  title = "Kaplan–Meier Curves by Treatment Group",
  xlab = "Time (days)", ylab = "S(t)", surv.median.line = "none",
  tables.theme = theme_cleantable()
))
dev.off()

# KM by sex
png("outputs/plots/km_by_sex.png", width = 900, height = 700, res = 120)
print(ggsurvplot(
  km_sex, data = rats_clean, conf.int = TRUE, risk.table = TRUE,
  pval = TRUE, palette = c("#A23B72", "#3B7080"),
  legend.labs = c("Female", "Male"), legend.title = "Sex",
  ggtheme = theme_minimal(),
  title = "Kaplan–Meier Curves by Sex",
  xlab = "Time (days)", ylab = "S(t)",
  tables.theme = theme_cleantable()
))
dev.off()

# Nelson-Aalen cumulative hazard overall
png("outputs/plots/nelson_aalen_overall.png", width = 900, height = 600, res = 120)
print(ggsurvplot(
  na_overall, data = rats_clean, fun = "cumhaz", conf.int = TRUE,
  palette = "#2E86AB", ggtheme = theme_minimal(),
  title = "Nelson–Aalen Cumulative Hazard",
  xlab = "Time (days)", ylab = "H(t)"
))
dev.off()

# Nelson-Aalen by treatment
png("outputs/plots/nelson_aalen_by_treatment.png", width = 900, height = 600, res = 120)
print(ggsurvplot(
  na_rx, data = rats_clean, fun = "cumhaz", conf.int = TRUE,
  palette = c("#2E86AB", "#F18F01"),
  legend.labs = c("Control", "Treated"), legend.title = "Group",
  ggtheme = theme_minimal(),
  title = "Cumulative Hazard by Treatment Group",
  xlab = "Time (days)", ylab = "H(t)"
))
dev.off()

# Smoothed hazard vs Weibull
png("outputs/plots/hazard_na_vs_weibull.png", width = 900, height = 600, res = 120)
plot(na_data$time, na_data$smooth_haz,
     type = "l", col = "#2E86AB", lwd = 2,
     xlab = "Time (days)", ylab = "h(t)",
     main = "Nelson–Aalen vs Weibull Hazard")
lines(weibull_h$time, weibull_h$est,
      col = "#F18F01", lwd = 2, lty = 2)
legend("topleft",
       legend = c("Nelson–Aalen (smoothed)", "Weibull (parametric)"),
       col = c("#2E86AB", "#F18F01"), lwd = 2, lty = c(1, 2), bty = "n")
grid()
dev.off()

# Forest plot
png("outputs/plots/forest_plot.png", width = 900, height = 500, res = 120)
print(ggforest(cox_multi, data = rats_clean))
dev.off()

# Schoenfeld residuals
png("outputs/plots/schoenfeld_residuals.png", width = 900, height = 500, res = 120)
par(mfrow = c(1, 2))
plot(ph_test, var = 1, main = "Schoenfeld — Treatment",
     ylab = "Beta(t)", xlab = "Time (days)", col = "#F18F01", lwd = 2)
abline(h = coef(cox_multi)[1], col = "red", lty = 2, lwd = 2)
plot(ph_test, var = 2, main = "Schoenfeld — Sex",
     ylab = "Beta(t)", xlab = "Time (days)", col = "#A23B72", lwd = 2)
abline(h = coef(cox_multi)[2], col = "red", lty = 2, lwd = 2)
par(mfrow = c(1, 1))
dev.off()

# dfbeta influential observations
png("outputs/plots/dfbeta_residuals.png", width = 900, height = 500, res = 120)
par(mfrow = c(1, 2))
plot(dfb[, 1], ylab = "dfbeta (treatment)",
     main = "Influence on Treatment Coefficient",
     pch = 16, cex = 0.6, col = "#2E86AB")
abline(h = 0, lty = 2, col = "red")
plot(dfb[, 2], ylab = "dfbeta (sex)",
     main = "Influence on Sex Coefficient",
     pch = 16, cex = 0.6, col = "#A23B72")
abline(h = 0, lty = 2, col = "red")
par(mfrow = c(1, 1))
dev.off()

# Martingale and deviance residuals
png("outputs/plots/martingale_deviance.png", width = 900, height = 500, res = 120)
par(mfrow = c(1, 2))
plot(rats_clean$time, mart_null,
     pch = 16, cex = 0.5, col = "#2E86AB",
     xlab = "Time (days)", ylab = "Martingale Residual",
     main = "Martingale Residuals vs Time")
lines(lowess(rats_clean$time, mart_null), col = "red", lwd = 2)
abline(h = 0, lty = 2)
plot(dev_resid, pch = 16, cex = 0.5, col = "#43A047",
     ylab = "Deviance Residual", main = "Deviance Residuals")
abline(h = c(-2, 0, 2), lty = c(2, 1, 2), col = c("red", "black", "red"))
par(mfrow = c(1, 1))
dev.off()

# KM vs Weibull overlay
png("outputs/plots/km_vs_weibull.png", width = 900, height = 600, res = 120)
plot(km_overall, col = "black", lwd = 2.5, conf.int = FALSE,
     xlab = "Time (days)", ylab = "S(t)",
     main = "KM Curve vs Best Parametric Model (Weibull)", ylim = c(0, 1))
lines(wb_surv$time, wb_surv$est, col = "#F18F01", lwd = 2, lty = 2)
legend("bottomleft",
       legend = c("Kaplan–Meier", "Weibull (parametric)"),
       col = c("black", "#F18F01"), lwd = 2, lty = c(1, 2), bty = "n")
grid()
dev.off()

# ── TABLES ─────────────────────────────────────────────────────────────────

write.csv(data.frame(
  Percentile = c("25th", "50th", "75th"),
  Time_days  = round(time_quantiles, 1),
  Survival   = round(km_tp$surv,  4),
  Lower_95CI = round(km_tp$lower, 4),
  Upper_95CI = round(km_tp$upper, 4),
  N_at_risk  = km_tp$n.risk
), "outputs/tables/survival_at_timepoints.csv", row.names = FALSE)

write.csv(data.frame(
  Method      = c("Log-Rank", "Wilcoxon"),
  Treatment_p = round(c(pval_lr_trt, pval_wc_trt), 4),
  Sex_p       = round(c(pval_lr_sex, pval_wc_sex), 4)
), "outputs/tables/hypothesis_tests.csv", row.names = FALSE)

write.csv(
  tidy(cox_multi, exponentiate = TRUE, conf.int = TRUE) %>%
    mutate(across(where(is.numeric), ~round(.x, 3))),
  "outputs/tables/cox_multivariable.csv", row.names = FALSE
)

write.csv(
  tidy(cox_strat, exponentiate = TRUE, conf.int = TRUE) %>%
    mutate(across(where(is.numeric), ~round(.x, 3))),
  "outputs/tables/cox_stratified.csv", row.names = FALSE
)

write.csv(
  tidy(cox_td, exponentiate = TRUE, conf.int = TRUE) %>%
    mutate(across(where(is.numeric), ~round(.x, 3))),
  "outputs/tables/cox_time_dependent.csv", row.names = FALSE
)

write.csv(
  as.data.frame(ph_test$table) %>%
    mutate(Variable = rownames(.), across(where(is.numeric), ~round(.x, 4))),
  "outputs/tables/ph_schoenfeld_test.csv", row.names = FALSE
)

cat("All plots saved to outputs/plots/\n")
cat("All tables saved to outputs/tables/\n")
cat("\nPlots saved:\n")
cat(paste(" -", list.files("outputs/plots")), sep = "\n")
cat("\nTables saved:\n")
cat(paste(" -", list.files("outputs/tables")), sep = "\n")
```

